---
title: "Clinical ADaM Pipeline"
subtitle: "CDISC Pilot Study — Xanomeline vs Placebo"
author: "Jeremy"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)

library(admiral)
library(pharmaversesdtm)
library(dplyr)
library(lubridate)
library(Tplyr)
library(gt)
library(survival)
library(ggsurvfit)
```

# Introduction and Overview

This lab demonstrates a complete ADaM data pipeline built using the pharmaverse R ecosystem. Source data is drawn from the CDISC pilot SDTM dataset, which represents a real Phase III clinical trial comparing Zanomeline (low and high dose) against Placebo in Alzheimer's disease patients.

The pipeline follows the same standards used in FDA regulatory submissions - CDISC SDTM as input, ADaM datasets as analysis-ready output, and TLFs (Tables, Listings, Figures) as the final deliverable.

---

# Data sources

All source data comes from the 'pharmaversedtm' R package, which contains CDISC-compliant SDTM test datasets maintained by the pharmaverse community.

The following SDTM domains are used:

Domain | Description              | Rows  |
|------|--------------------------|-------|
DM      Demographics                One per patient
EX      Exposure (dosing records)   Many per patient
AE      Adverse Events              Many per patient
DS      Disposition                 One per patient

---

# ADaM Datasets

## ADSL - Subject Level Analysis Dataset

ADSL is the foundation of every ADaM package. It contains exactly one row per patient and captures baseline characteristics, treatment, assignment, and population flags. Every downstream dataset joins back to ADSL using 'USUBJID' as the key.

# Loading in ADSL
```{r load-adsl}
adsl <- readRDS("data/adam/adsl.rds")
```

# Population Summary
```{r adsl-summary}
adsl |>
  count(ARM, SAFFL) |>
  gt() |>
  tab_header(title = "Patient Counts by Treatment Arm and Safety Flag") |>
  cols_label(
    ARM = "Treatment Arm",
    SAFFL = "Safety Population",
    n = "N"
  )
```
### Baseline Demographics
```{r adsl-demographics}
adsl |>
  filter(SAFFL == "Y") |>
  group_by(ARM) |>
  summarise(
    N = n()
    Mean_Age = round(mean(AGE, na.rm = TRUE), 1),
    SD_Age = 
  )
